\documentclass[11pt, a4paper]{article}

\usepackage[margin=1.1in]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{parskip}
\usepackage{titlesec}
\usepackage{xcolor}

\titleformat{\section}{\large\bfseries}{}{0em}{}[\titlerule]
\titleformat{\subsection}{\normalsize\bfseries}{}{0em}{}

\hypersetup{colorlinks=true, urlcolor=blue, linkcolor=black, citecolor=black}

\title{\textbf{Does Mamba Actually Help in Hybrid CNN Models?}\\[0.4em]
\large A Controlled Study of Mamba Against MLP and Attention in Tiny Vision Models}
\author{}
\date{}

\begin{document}
\maketitle

% ─────────────────────────────────────────────────────────────
\section{The Problem}

Mamba has been proposed as an efficient alternative to self-attention for sequence
modeling in vision. The argument is straightforward: linear complexity in sequence
length, no need for positional encodings, and competitive accuracy. Hybrid CNN-Mamba
architectures have been built on this premise — but the premise itself has not been
tested carefully.

Two questions have gone unanswered. First, does Mamba actually outperform a
parameter-matched MLP, or is the CNN stem doing all the work regardless of what
sequence block sits on top? Second, does Mamba genuinely beat attention in vision
tasks, or does attention recover its advantage when given enough task complexity?
Without answers to both, there is no principled reason to choose Mamba over simpler
or more established alternatives.

% ─────────────────────────────────────────────────────────────
\section{Research Objective}

We study \textbf{when and by how much Mamba outperforms both a parameter-matched MLP
and multi-head self-attention} as the sequence processing block in a hybrid CNN model.
The two variables we control are task complexity and sequence length $L$.

Our hypothesis is that Mamba's selective gating mechanism only activates useful
recurrent structure when (a) the task requires global spatial reasoning beyond CNN
receptive fields, and (b) the sequence is long enough for the recurrence to matter.
Below those thresholds, Mamba should be equivalent to an MLP. Against attention
specifically, Mamba should hold an advantage at longer sequences since attention, without
positional encodings, struggles as the token count grows in tiny models.

% ─────────────────────────────────────────────────────────────
\section{Setup}

All five model variants share an identical CNN stem and classification head.
Only the sequence processing block differs:

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Variant} & \textbf{Sequence Block} \\
\midrule
\texttt{pure\_cnn}       & None — CNN $\to$ Global Avg Pool $\to$ Head \\
\texttt{cnn\_mlp}        & 2-layer FFN, $4\times$ expansion (parameter-matched) \\
\texttt{cnn\_mamba\_uni} & Unidirectional Mamba (S6) \\
\texttt{cnn\_mamba\_bi}  & Bidirectional Mamba (S6) \\
\texttt{cnn\_attn}       & Multi-head self-attention + FFN \\
\bottomrule
\end{tabular}
\end{center}

Sequence length $L$ is controlled by the number of pooling stages in the CNN stem
(\texttt{n\_pool}). All datasets are resized to $96\times96$, giving identical $L$
values across datasets:

\begin{center}
\begin{tabular}{ccc}
\toprule
\texttt{n\_pool} & Token grid & $L$ \\
\midrule
4 & $6\times6$   & 36  \\
3 & $12\times12$ & 144 \\
2 & $24\times24$ & 576 \\
\bottomrule
\end{tabular}
\end{center}

Three datasets span the task-complexity axis:

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Dataset} & \textbf{Classes} & \textbf{Train size} & \textbf{Task type} \\
\midrule
DTD (Describable Textures) & 47  & 1,880  & Pure local texture \\
STL-10                     & 10  & 4,500  & Shape + texture mix \\
CIFAR-100                  & 100 & 45,000 & Fine-grained objects \\
\bottomrule
\end{tabular}
\end{center}

Each configuration is trained with 3 random seeds (0, 42, 99) and reported as
mean $\pm$ std. The train/val/test split is fixed across seeds (\texttt{split\_seed=42}),
so reported variance reflects only weight initialization and batch ordering.

% ─────────────────────────────────────────────────────────────
\section{Training Status}

DTD and STL-10 are \textbf{fully complete} — all 5 models $\times$ 3 sequence lengths
$\times$ 3 seeds. CIFAR-100 is partially complete. The missing runs are:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Run} & \textbf{Status} & \textbf{Remaining compute} \\
\midrule
CIFAR-100, Mamba (Uni), $L=576$ & Not started & $\sim$8--10 GPU hours \\
CIFAR-100, Mamba (Bi),  $L=576$ & Not started & $\sim$8--10 GPU hours \\
CIFAR-100, Attention,   $L=36$  & 1 of 3 seeds & $\sim$4 GPU hours \\
\bottomrule
\end{tabular}
\end{center}

The two missing Mamba runs at $L=576$ on CIFAR-100 are the most critical gap —
they would complete the long-sequence column for the hardest task, precisely where
Mamba is expected to show its strongest advantage over attention.

% ─────────────────────────────────────────────────────────────
\section{Results}

\subsection{DTD — Pure Texture}

\begin{center}
\begin{tabular}{lccc}
\toprule
\textbf{Model} & $L=36$ & $L=144$ & $L=576$ \\
\midrule
Pure CNN          & 26.44 $\pm$ 0.16 & 24.20 $\pm$ 0.50 & 22.52 $\pm$ 0.46 \\
CNN + MLP         & 25.66 $\pm$ 0.81 & 25.39 $\pm$ 0.55 & 25.66 $\pm$ 0.13 \\
CNN + Mamba (Uni) & 23.81 $\pm$ 0.37 & 24.96 $\pm$ 1.23 & 24.95 $\pm$ 1.59 \\
CNN + Mamba (Bi)  & 22.68 $\pm$ 0.71 & 24.73 $\pm$ 0.92 & 25.02 $\pm$ 1.45 \\
CNN + Attention   & \textbf{25.48 $\pm$ 0.43} & \textbf{26.08 $\pm$ 0.16} & 25.05 $\pm$ 0.66 \\
\bottomrule
\end{tabular}
\end{center}

All five models land within 3\% of each other regardless of sequence length.
Neither Mamba nor attention offers any advantage over the MLP, and the pure CNN
essentially ties the field. This is the expected null result — for texture tasks,
CNN features already carry all necessary information and the sequence block
is irrelevant regardless of its type.

\subsection{STL-10 — Mixed Complexity}

\begin{center}
\begin{tabular}{lccc}
\toprule
\textbf{Model} & $L=36$ & $L=144$ & $L=576$ \\
\midrule
Pure CNN          & 65.83 $\pm$ 0.78 & 58.50 $\pm$ 0.44 & 52.40 $\pm$ 0.80 \\
CNN + MLP         & 69.34 $\pm$ 0.11 & 64.65 $\pm$ 0.39 & 63.40 $\pm$ 0.73 \\
CNN + Mamba (Uni) & 71.80 $\pm$ 0.08 & 71.97 $\pm$ 0.43 & \textbf{72.67 $\pm$ 1.14} \\
CNN + Mamba (Bi)  & 71.02 $\pm$ 0.66 & 71.48 $\pm$ 0.34 & 72.14 $\pm$ 0.26 \\
CNN + Attention   & 69.86 $\pm$ 0.73 & 66.57 $\pm$ 0.33 & 66.97 $\pm$ 0.58 \\
\bottomrule
\end{tabular}
\end{center}

This is where the results become interesting. Mamba outperforms the MLP by $\sim$3\%
and, more importantly, \textbf{outperforms attention by up to 5.7\%} — a gap that
grows with sequence length. At $L=36$, Mamba leads attention by 1.9\%. By $L=576$,
that gap widens to 5.7\%. Attention degrades from 69.86\% at $L=36$ down to 66.97\%
at $L=576$, while Mamba moves in the opposite direction (71.80\% $\to$ 72.67\%).

The pattern is consistent: attention without positional encodings cannot maintain
its representational quality as the token count grows. Mamba's recurrent structure
handles longer sequences without degradation.

\subsection{CIFAR-100 — Fine-Grained Recognition (Partial)}

\begin{center}
\begin{tabular}{lccc}
\toprule
\textbf{Model} & $L=36$ & $L=144$ & $L=576$ \\
\midrule
Pure CNN          & 38.95 $\pm$ 0.42 & 28.59 $\pm$ 0.25 & 26.67 $\pm$ 0.31 \\
CNN + MLP         & 50.43 $\pm$ 0.15 & 44.35 $\pm$ 0.63 & 43.16 $\pm$ 0.20 \\
CNN + Mamba (Uni) & 54.05 $\pm$ 0.50 & 53.65 $\pm$ 0.36 & \textit{not trained} \\
CNN + Mamba (Bi)  & 53.98 $\pm$ 0.83 & 53.47 $\pm$ 0.39 & \textit{not trained} \\
CNN + Attention   & $59.70^{\,\dagger}$ & 50.25 $\pm$ 0.26 & 52.51 $\pm$ 0.37 \\
\bottomrule
\multicolumn{4}{l}{\footnotesize $\dagger$ Single seed only — treat as indicative, not conclusive.}
\end{tabular}
\end{center}

At $L=144$, Mamba outperforms attention by 3.4\%, consistent with STL-10.
The single-seed attention result at $L=36$ (59.70\%) suggests attention may recover
at very short sequences on harder tasks — but with only one seed, this cannot be
treated as a firm result. The missing $L=576$ column for Mamba is the critical open
question: attention sits at 52.51\% there, and based on the STL-10 trend, Mamba
would be expected to match or exceed that.

% ─────────────────────────────────────────────────────────────
\section{Findings}

\textbf{Mamba vs.\ MLP.} On texture tasks, Mamba and MLP are equivalent —
the CNN stem saturates all useful features before the sequence block.
On object recognition (STL-10, CIFAR-100), Mamba outperforms the MLP by 3--4\%,
and the gap is consistent across seeds and datasets. The selective gating mechanism
is genuinely doing useful work for tasks that require global spatial reasoning.

\textbf{Mamba vs.\ Attention.} This is the stronger finding.
On STL-10, Mamba outperforms attention at every sequence length, with the margin
growing from 1.9\% at $L=36$ to 5.7\% at $L=576$. Attention degrades as $L$ grows;
Mamba does not. On CIFAR-100, Mamba leads attention at $L=144$ by 3.4\%.
The practical implication is direct: for medium-to-long token sequences in tiny
hybrid models, Mamba is the better choice over attention — not just in compute
efficiency but in accuracy.

\textbf{Bidirectionality does not matter.}
Unidirectional and bidirectional Mamba perform within noise margin on every dataset
and every sequence length. For spatially tokenized image features, the backward
scan adds nothing. This is a useful design guideline: the simpler unidirectional
variant is sufficient.

\textbf{Where the story is incomplete.}
The $L=576$ column for Mamba on CIFAR-100 is missing. This is where the Mamba
vs.\ attention comparison on the hardest task at the longest sequence would be
decided. The partial evidence points toward a Mamba advantage, but it cannot
be stated conclusively without those runs.

% ─────────────────────────────────────────────────────────────
\section{Compute Request}

To complete the study, we need approximately \textbf{22--24 GPU hours} on a single
A100 (or equivalent) to train the three remaining configurations:

\begin{itemize}
    \item Mamba (Uni) and Mamba (Bi) at $L=576$ on CIFAR-100 — 3 seeds each
    \item Attention at $L=36$ on CIFAR-100 — 2 remaining seeds
\end{itemize}

These runs would close the ablation matrix and allow a definitive comparison of
Mamba against attention at long sequences on a high-complexity task — the exact
regime where the outcome is most contested and most useful.

% ─────────────────────────────────────────────────────────────
\section{Contributions}

\begin{enumerate}
    \item Controlled evidence that Mamba outperforms attention in tiny hybrid vision
    models at medium-to-long sequence lengths, with a growing margin as $L$ increases.
    \item A task-complexity threshold: below a certain complexity (pure texture),
    neither Mamba nor attention offers anything over an MLP.
    \item Evidence that bidirectional Mamba is unnecessary for spatial image tokens —
    unidirectional Mamba is sufficient and cheaper.
    \item A reusable benchmark framework for evaluating future SSM architectures
    at sub-1MB scale across a controlled complexity axis.
\end{enumerate}

\end{document}
